---
title: "Affymetrix"
author: "Mark Dunning"
date: "20/11/2015"
output: html_document
---


Walkthrough of example dataset pointing out various QA steps as they occur

# Technology overview



# Example analysis in R

## The estrogen dataset

The data for this section are described the `estrogen` data package

```{r echo=FALSE,cache=TRUE}
datadir <- system.file("extdata", package="estrogen")
dir.create("estrogen")
sapply(list.files(datadir), function(x) file.copy(paste(datadir,x,sep="/"),"estrogen"))
```


Raw data for Affymetrix arrays are usually stored in `.cel` format. The cel files for the example experiement are stored in the `estrogen` directory


We read the targets file
```{r message=FALSE}
library(affy)
targetsFile <- "estrogen/estrogen.txt"
pd <- read.AnnotatedDataFrame(targetsFile,header=TRUE,sep="",row.names=1)
pData(pd)
```


```{r,message=FALSE}

raw <-ReadAffy(celfile.path = "estrogen", filenames=rownames(pData(pd)),phenoData = pd)
raw
```

1) What type of Affy array is this
2) How many features?
3) How many samples?

We can look at the images of the array surface. Ocassionally this can reveal [http://plmimagegallery.bmbolstad.com/](interesting / unusual artefacts)

If you want to look at the array image, we can use the following code

```{r}
bad <- ReadAffy(celfile.path = "estrogen/",filenames="bad.cel")
image(bad)
```

Try out some images for this dataset

```{r eval=FALSE}
par(mfrow=c(2,4))
image(raw[,1])
image(raw[,2])
image(raw[,3])
image(raw[,4])
image(raw[,5])
image(raw[,6])
image(raw[,7])
image(raw[,8])
```

- *Do you see any problems?

The first diagnostic plot we will meet is the "boxplot"

```{r echo=FALSE}
df <- data.frame(x=rnorm(20),y=rnorm(20,2,2))
boxplot(df)
```

In Bioconductor, you will usually find that package authors have created shortcuts to allow complicated data types to be visualised with common functions.

```{r}
boxplot(raw,col="red",las=2)
```

* What do you notice from this plot?*

## Perfect-Match and Mismatch probes


```{r}
par(mfrow=c(2,1))
hist(log2(pm(raw[,1])),breaks=100,col="steelblue",main="PM",xlim=c(4,14))
hist(log2(mm(raw[,1])),breaks=100,col="steelblue",main="MM",xlim=c(4,14))

```

## Probe-level Linear Models

```{r}
library(affyPLM)
plmset <- fitPLM(raw)
NUSE(plmset,las=2)
RLE(plmset,las=2)
```


## Normalisation


- We want to be observing *biological* and not *technical* variation
- We wouldn't expect such wholesale changes on a per-sample basis
- Easy option would to scale values for each array to median level
```{r echo=FALSE,warning=FALSE,fig.height=4,fig.width=8}

boxplot(raw)

x <- raw
         tmp <- unlist(indexProbes(x, which="both"))
         tmp <- tmp[seq(1, length(tmp), len = 5000)]
         df <- data.frame(log2(intensity(x)[tmp, ]))
        med <- median(as.matrix(df))

abline(h=med,lty=2,col="red")
```

## Simple scaling

- Genes on array 2 are on average `r median(df[,2]) - med` higher than the global median, so subtract `r median(df[,2]) - med` from each gene
- Genes on array 8 are on average `r abs(median(df[,8]) - med)` lower than the global median, so add `r abs(median(df[,8]) - med)` to each gene
- etc


## Non-linear effects

The *MA-plot* is commonly-used in microarray analysis to 

```{r}
mva.pairs(pm(raw)[,1:2])
```


## Quantile normalisation

This is arguably the most-popular normalisation method available. Consider the following matrix of values to be normalised

```{r echo=FALSE}
df <- data.frame(Array1 = c(1,3,9,2,4), Array2 = c(3,4,2,1,9), Array3 = c(9,1,5,7,6))
rownames(df) <- LETTERS[1:nrow(df)]
df
```
Genes A, B, C, D and E measured on three arrays



Determine ranks of each column
```{r echo=FALSE}
df
```

```{r echo=FALSE}
rks <- apply(df, 2, function(x) paste("Rank",rank(x,ties.method="min"),sep=""))
rks
```



Sort each column Largest...Smallest

Original data
```{r echo=FALSE}
df
```
***
Sorted data
```{r echo=FALSE}
apply(df, 2,sort)
```
Then calculate target distribution by averaging the sorted rows
```{r echo=FALSE} 
target <- round(rowMeans(apply(df, 2,sort)),3)
names(target) <- paste("Rank", 1:length(target),sep="")
target
```




Go back to the rank matrix

```{r echo=FALSE}
rks

```
Substitue with values from the target distribution
```{r echo=FALSE}
target
```

```{r echo=FALSE}
rks[,1] <- gsub("Rank1",target["Rank1"],rks[,1])
rks
```



Go back to the rank matrix

```{r echo=FALSE}
rks

```

Substitue with values from the target distribution

```{r echo=FALSE}
target
```

```{r echo=FALSE}
rks[,1] <- gsub("Rank2",target["Rank2"],rks[,1])
rks
```




Go back to the rank matrix

```{r echo=FALSE}
rks

```
Substitue with values from the target distribution
```{r echo=FALSE}
target
```

```{r echo=FALSE}
rks[,1] <- gsub("Rank3",target["Rank3"],rks[,1])
rks
```


Go back to the rank matrix

```{r echo=FALSE}
rks

```
Substitue with values from the target distribution
```{r echo=FALSE}
target
```

```{r echo=FALSE}
rks[,1] <- gsub("Rank4",target["Rank4"],rks[,1])
rks
```


Go back to the rank matrix

```{r echo=FALSE}
rks

```
Substitue with values from the target distribution
```{r echo=FALSE}
target
```

```{r echo=FALSE}
rks[,1] <- gsub("Rank5",target["Rank5"],rks[,1])
rks
```


We then repeat to get the normalized matrix
```{r echo=FALSE}
for(i in 1:3){
rks[,i] <- gsub("Rank1",target["Rank1"],rks[,i])
rks[,i] <- gsub("Rank2",target["Rank2"],rks[,i])
rks[,i] <- gsub("Rank3",target["Rank3"],rks[,i])
rks[,i] <- gsub("Rank4",target["Rank4"],rks[,i])
rks[,i] <- gsub("Rank5",target["Rank5"],rks[,i])
}
rks <- as.data.frame(rks)
rownames(rks) <- rownames(df)
```

Original data
```{r echo=FALSE}
df
```
Normalised data

```{r echo=FALSE}
rks
```


##Final Code

```{r}
df <- data.frame(Array1 = c(1,3,9,2,4), 
                 Array2 = c(3,4,2,1,9), Array3 = c(9,1,5,7,6))
rownames(df) <- LETTERS[1:nrow(df)]
rks <- apply(df, 2, function(x) paste("Rank",
      rank(x,ties.method="min"),sep=""))
target <- round(rowMeans(apply(df, 2,sort)),3)
names(target) <- paste("Rank", 1:length(target),sep="")
for(i in 1:ncol(df)){
  for(nm in names(target)){
    rks[,i] <- gsub(nm,target[nm],rks[,i])  
      }
}
norm <- as.data.frame(rks)
```


##Effect of quantile normalisation

Caveats

- Distributions of samples are expected to be the same
- Majority of genes do not change between groups

## Other normalisation procedures

- loess or splines

## Normalising the Estrogen dataset

rma incorporates quantile normalisation, along with a background correction

The result is an `ExpressionSet`, which is a ubiquitous object in Biooconductor for storing high-throughput data

```{r}
eset <- rma(raw)
eset
```

```{r}
head(exprs(eset))
```

```{r}
head(pData(eset))
```


```{r}
boxplot(exprs(eset))
```

```{r}
mva.pairs(exprs(eset)[,1:4])
```

## Automated QC

```{r}
library(arrayQualityMetrics)
arrayQualityMetrics(eset)
```


## Summary

- Affy data come in `.cel` files that can be imported with the `affy` package
- 


