---
title: "Clustering and classification"
author: "Mark Dunning"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---

```{r echo=FALSE,message=FALSE}
library(GEOquery)
library(limma)
```

We will re-use the colon cancer data in GSE33126

## Importing the data

```{r cache=TRUE}
library(GEOquery)
url <- "ftp://ftp.ncbi.nih.gov/pub/geo/DATA/SeriesMatrix/GSE33126/"
filenm <- "GSE33126_series_matrix.txt.gz"
if(!file.exists("GSE33126_series_matrix.txt.gz")) download.file(paste(url, filenm, sep=""), destfile=filenm)
colonData <- getGEO(filename=filenm)
colonData
```

```{r}
exprs (colonData) <- log2 (exprs(colonData))
SampleGroup <- pData(colonData)$source_name_ch1
Patient <- pData(colonData)$characteristics_ch1.1
```

## Filtering the data

A first step in the analysis of microarray data is often to remove any uniformative probes.
We can do this because typically only 50% of probes genes will be expressed, and even fewer
than this will be differentially expressed. Including such non-informative genes in visualisa-
tion will obscure any biological differences that exist. The genefilter package contains a suite
of tools for the filtering of microarray data. The varFilter function allows probes with low-
variance to be removed from the dataset. The metric using to decide which probes to remove
is the Inter-Quartile Range (IQR), and by default half of the probes are removed. Both the
function used to do the filtering, and cut-off can be specified by the user.


```{r}
library (genefilter)
dim (colonData)
varFiltered <- varFilter (colonData)
dim (varFiltered)
nrow (colonData) / nrow (varFiltered)
```

## Calculating a distance matrix

The first step towards clustering the data is to construct a distance matrix. Each entry in
this matrix is the pairwise distance between two samples. Note that for expression data we have to transpose the standard ExpressionSet format, as clustering is designed to work on
rows rather than columns. The standard function to make a distance matrix in R is dist
which uses the euclidean metric. As the data entries in a distance matrix are symmetrical, it
has a different representation to the default matrix (i.e. values are not repeated unnecessarily), and clustering algorithms are able to accept this representation as input.

```{r}
euc.dist <- dist (t(exprs(varFiltered)))
euc.dist
```

For gene-expression data, it is common to use correlation as a distance metric rather than
the Euclidean. You should make sure that you know the difference between the two metrics.
The cor function can be used to calculate the correlation of columns in a matrix. Each row
(or column) in the resulting matrix is the correlation of that sample with all other samples
in the dataset. The matrix is symmetrical and we can transform this into a distance matrix
by first subtracting 1 from the correlation matrix. Hence, samples with a higher correlation
have a smaller ’distance’.


```{r}
corMat <- cor(exprs(varFiltered))
corMat
cor.dist <- as.dist(1 - corMat)
```

## Hierachical clustering

Hierachical clustering is the method by which samples with similar profiles are grouped to-
gether, based on their distances. The method for grouping samples can be specified, with the
default being to use complete linkage. Different linkage methods can affect the properties of
the clustering output. e.g. the ’ward’ method tends to produce smaller, compact clusters. A
popular way of displaying the results of clustering is a dendrogram, which arranges the sam-
ples on the x-axis and shows the distances between samples on the y-axis. It is important to
note that the distance of samples on the x-axis has no meaning. The fact that two samples
are displayed next to each other, does not m


```{r}
par(mfrow = c (1 , 2))
clust.euclid = hclust(euc.dist)
clust.cor = hclust (cor.dist)
par (mfrow = c (1 , 2))
plot(clust.euclid , label = SampleGroup )
plot(clust.cor , label = SampleGroup )
```

## Extracting data from the clustering

If we want to interpret the data presented in a clustering analysis, we need a way of extract-
ing which samples are grouped together, or to determine the optimal grouping of samples.
One intuitive way of assigning groups it to ’cut’ the dendrogram at a particular height on
the y-axis. We can do this manually on the plot, or use the cutree function to return the
labels of samples that are belong to the same group when the dendrogram is cut at the specified height, h. Alternatively, we can specify how many groups, k, that we want to create.

```{r}
library (cluster)
par (mfrow = c(1 , 1))
plot(clust.cor)
abline (h = 0.12, col = " red ")
cutree (clust.cor , h = 0.12)
cutree (clust.cor , k = 2)
table (cutree(clust.cor , k = 3) , SampleGroup)
```

A Silhouette plot can be used to choose the optimal number of clusters. For each sample, we
calculate a value that quantifies how well it ’fits’ the cluster that it has been assigned to. If
the value is around 1, then the sample closely fits other samples in the same cluster. How-
ever, if the value is around 0 the sample could belong to another cluster. In the silhouette
plot, the values for each cluster are plotted together and ordered from largest to smallest.
The number of samples belonging to each group is also displayed.

```{r}
par (mfrow = c (2 , 2))
plot (silhouette(cutree(clust.cor, k=2),cor.dist),
      col="red",main=paste("k=",2))
plot (silhouette(cutree(clust.cor, k=3),cor.dist),
      col="red",main=paste("k=",3))
plot (silhouette(cutree(clust.cor, k=4),cor.dist),
      col="red",main=paste("k=",4))
plot (silhouette(cutree(clust.cor, k=5),cor.dist),
      col="red",main=paste("k=",5))
```


If we have prior knowledge of how many clusters to expect, we could run the clustering in a
supervised manner. The ***Partition Around Medioids*** method can be used to group samples
into k clusters.

```{r}
pam.clus <- pam (euc.dist , k = 2)
clusplot (pam.clus)
pam.clus$clustering
table(pam.clus$clustering , SampleGroup)
```

## Producing a heatmap

A heatmap is often used to visualise differences between samples. Each row represents a
gene and each column is an array and coloured cells indicate the expression levels of genes.
Both samples and genes with similar expression profile are clustered together. By default,
euclidean distances are used with complete linkage clustering.
Drawing a heatmap in R uses a lot of memory and can take a long time, therefore reducing the amount of data to be plotted is usually recommended. Including too many non-
informative genes can also make it difficult to spot patterns. Typically, data are filtered to
include the genes which tell us the most about the biological variation. In an un-supervised
setting, the selection of such genes is done without using prior knowledge about the sample
groupings.

```{r}
IQRs = apply (exprs(varFiltered) , 1 , IQR )
highVarGenes = order (IQRs, decreasing = T )[1:100]
Symbols <- as.character(fData(colonData)$Symbol[highVarGenes])
heatmap (as.matrix(exprs(varFiltered)[highVarGenes, ]),
         labCol = SampleGroup , labRow = Symbols)

```

The default options for the heatmap are to cluster both the genes (rows) and samples (columns).
However, sometimes we might want to specify a particular order. For example, we might
want to order the columns according to sample groups. We can do this by re-ordering the
input matrix manually and setting the `Colv` argument to `NA`. This tells the heatmap function not be
cluster the columns. It choosing this option, we need to be careful that any column colourings or labels are set to the same ordering. Alternatively, a pre-calculated dendrogram could be used.

```{r}
clus.ward <- hclust (cor.dist , method = "ward")
heatmap (as.matrix(exprs(varFiltered)[highVarGenes, ]) ,
         Colv = as.dendrogram(clus.ward) , labCol = SampleGroup )

```

## Customising the heatmap

The heatmap function can be customised in many ways to make the output more informa-
tive. For example, the labRow and labCol parameters can be used to give labels to the rows
(genes) and columns (sample) of the heatmap. Similarly, `ColSideColors` and `RowSideColors` give coloured labels, often used to indicate different groups which are know in advance.
See the help page for heatmap for more details.

```{r}
labs <- as.factor(SampleGroup)
levels(labs) <- c ("yellow" , "blue")
heatmap(as.matrix(exprs(varFiltered)[highVarGenes, ]) ,
        labCol = Patient, ColSideColors = as.character(labs),
        labRow = Symbols)
```

The colours used to display the gene expression values can also be modified. For this, we
can use the RColorBrewer package which has functions for creating pre-defined palettes. The
function display.brewer.all can be used to display the palettes available through this
package.

```{r}
library (RColorBrewer)
display.brewer.all()
hmcol <- brewer.pal(11 , "RdBu")
heatmap (as.matrix(exprs(varFiltered)[highVarGenes, ]) ,
  ColSideColors = as.character(labs) , labRow = Symbols,
  col=hmcol)
```

You should avoid using the traditional red / green colour scheme as it may be difficult for
people with colour-blindness to interpret

A popular use for heatmaps is to take an existing gene list (e.g. genes found to be significant
in a previous study, or genes belonging to a particular pathway) and produce an image of
how they cluster the data for exploratory purposes. This can be achieved by selecting
the correct rows in the data matrix.

Here we use the pre-built annotation package for the particular array platform (`illuminaHumanv3.db`). These packages contain mappings between the manufacturer identifiers, and more-common genomic identifiers. The `illuminaHumanv3PATH` mapping will take a set of illumina IDs and return the KEGG pathway for each. However, if we want the Illumina IDs for a particular KEGG pathway, we have to *reverse* the mapping. 

```{r}
library (illuminaHumanv3.db)
pathwayGenes <- unlist (mget("04110" , revmap(illuminaHumanv3PATH)))
pathwayGenes <- pathwayGenes [pathwayGenes %in% featureNames(varFiltered)]
symbols <- fData(varFiltered)[pathwayGenes , "Symbol"]
heatmap (as.matrix(exprs(varFiltered)[pathwayGenes , ]),
  ColSideColors = as.character (labs) , labCol = Patient ,
  labRow = symbols , col = hmcol )

```

Later on, we will show how other identifier mapping can be performed using so-called organism packages.

## Principal Components Analysis

Principal components analysis (PCA) is a data reduction technique that allows us to simplify multidimensional data sets to 2 or 3 dimensions for plotting purposes and identify which
factors explain the most variability in the data. We can use the prcomp function to perform
a PCA and we have to supply it with a distance matrix. The resulting object contains information on the proportion of variance that is ’explained’ by each component. Ideally, we want
to see that the majority of the variance is explained by the first 2 or 3 components, and that
these components are associated with a biological factor

```{r}
pca <- prcomp(exprs(varFiltered))
plot(pca)
summary(pca)
```

The ggplot2 package is convenient for plotting principal components results as it allows many
different covariates to be overlaid on the plot. See http://ggplot2.org/ for more informa-
tion on this package.

```{r}
library(ggplot2)
clusLabs <- cutree(clust.cor , k = 3)
pcRes <- data.frame(pca$rotation , SampleGroup , Sample = Patient)
ggplot (pcRes , aes(x = PC1 , y = PC2 , col = SampleGroup ,
                       label = Patient , pch = as.factor(clusLabs))) + geom_point () +
      geom_text(vjust=0,alpha=0.5)
  
```




# Classification

The Bioconductor project has a collection of example datasets. Often these are used as examples to illustrate a particular package or functionality, or to accompany the analysis presented in a publication. For example, several datasets are presented to accompany the genefu
package which has functions useful for the classification of breast cancer patients based on
expression profiles.
An experimental dataset can be installed and loaded as with any other Bioconductor package. The data itself is saved as an object in the package. You will need to see the documentation for the package to find out the relevant object name. The full list of datasets available
through Bioconductor can be found [here](http://bioconductor.org/packages/release/BiocViews.html#___ExperimentData)

```{r}
library(breastCancerVDX)
library(breastCancerTRANSBIG)
data(vdx)
data(transbig)
dim(vdx)
dim(transbig)
annotation(vdx)
annotation(transbig)
```

If we want any classifers to be reproducible and applicable to other datasets, it is sensible
to exclude probes that do not have sufficient annotation from the analysis. For this, we can
use the genefilter package as before. The nsFilter function performs this annotation-based
filtering as well as variance filtering. The output of the function includes details about how
many probes were removed at each stage of the filtering.

```{r}
library (genefilter)
vdx.filt <- nsFilter(vdx)
vdx.filt
vdx.filt <- vdx.filt[[1]]
```

Format the vdx data for pamr, and train a classifier to predict ER status. For extra
clarity in the results, it might be useful to rename the binary er status used in the data package
to something more descriptive.

```{r}
library(pamr)
dat <- exprs(vdx.filt)
gN <- as.character(fData(vdx.filt)$Gene.symbol)
gI <- featureNames (vdx.filt)
sI <- sampleNames (vdx.filt)
erStatus <- pData (vdx)$er
erStatus <- gsub (0 , "ER -" , erStatus )
erStatus <- gsub (1 , "ER +" , erStatus )

```

Fitting the model

```{r}
train.dat <- list ( x = dat , y = erStatus , genenames = gN ,
              geneid = gI , sampleid = sI )
model <- pamr.train(train.dat ,n.threshold = 100)
model
```

We can perform cross-validation using the pamr.cv function. Printing the output of this
function shows a table of how many genes were used at each threshold, and the number of
classification errors. Both these values need to be taken into account when choosing a suit-
able theshold. The `pamr.plotcv` function can assist with this by producing a diagnostic plot
which shows how the error changes with the number of genes. In the plot produced by this
function there are two panels; the top one shows the errors in the whole dataset and the bottom one considers each class separately. In each panel, the x axis corresponds to the thresh-
old (and number of genes at each threshold) whereas the y-axis is the number of misclassifications.

```{r}
model.cv <- pamr.cv(model , train.dat , nfold = 10)
model.cv
pamr.plotcv(model.cv)
```

In the following sections, feel free to experiment with different values of the threshold
(which we will call `Delta`)
The misclassifications can easily be visualised as a ’confusion table’. This simply tabulates
the classes assigned to each sample against the original label assigned to the sample. e.g.
Misclassifications are samples that we thought were ’ER+’ but have been assigned to the
’ER-’ group by the classifier, or ’ER-’ samples assigned as ’ER+’ by the classifier.

```{r}
Delta <- 8
pamr.confusion(model.cv , Delta)
```

A visual representation of the class separation can be obtained using the `pamr.plotcvprob`
function. For each sample there are two circles representing the probabilty of that sample
being classified ER- (red) or ER+ (green).

```{r}
pamr.plotcvprob(model , train.dat , Delta )
```

There are a couple of ways of extract the details of the genes that have been used in the
classifier. We can list their names using the pamr.listgenes function, which in our case
these are just returns the microarray probe names. We can however, use these IDs to query
the `featureData` stored with the original `vdx` object. We can also plot the expression values
for each gene, coloured according to the class label.


```{r, fig.width=12,fig.height=12}
pamr.listgenes(model , train.dat , Delta )
classifierGenes <- pamr.listgenes(model , train.dat , Delta )[,1]
pamr.geneplot(model , train.dat ,Delta)
```

***You may get an error message Error in plot.new(): Figure margins too large
when trying to produce the gene plot. If this occurs, try increasing the size of your plotting
window, or decrease the number of genes by decreasing the threshold. Alternatively, the fol-
lowing code will write the plots to a pdf.***

```{r}
pdf ("classifierProfiles.pdf")
for (i in 1: length (classifierGenes)) {
  Symbol <- fData(vdx.filt)[classifierGenes[i] , "Gene.symbol"]
  boxplot(exprs(vdx.filt)[classifierGenes[i], ] ~ erStatus ,
  main = Symbol )
}
dev.off()
```

Use the genes identified by the classifier to produce a heatmap to confirm that they
separate the samples as expected.

```{r}

symbols <- fData(vdx.filt)[classifierGenes , "Gene.symbol"]
heatmap(exprs(vdx.filt)[classifierGenes, ] , labRow = symbols )

```


## Testing the model

We can now test the classifier on an external dataset. We choose the transbig dataset for
simplicity as it was generated on the same microarray platform

```{r}
library (breastCancerTRANSBIG)
data (transbig)
pData (transbig)[1:4, ]
transbig.filt <- transbig [featureNames(vdx.filt) , ]

```

```{r}
predClass <- pamr.predict(model ,exprs(transbig.filt) ,Delta )
table (predClass, pData(transbig)$ er)
boxplot (pamr.predict(model , exprs(transbig.filt), Delta ,
                           type = "posterior" )[, 1] ~ pData(transbig)$er)
```

Make a heatmap of the transbig data using the genes involved in the vxd classifier

```{r}
erLab <- as.factor(pData(transbig)$er)
levels (erLab) <- c ("blue" , "yellow")

heatmap (exprs(transbig.filt)[classifierGenes , ] , labRow = symbols ,
  ColSideColors = as.character (erLab))
```


# Survival Analysis


An attractive feature of the vdx dataset is that it includes survival data for each breast can-
cer patient. We are not explicitly covering survival analysis in this course, but for your reference, here are the commands to create survival curves when patients are grouped by ER
status and tumour grade.

```{r}
library (survival)
par (mfrow = c (1 , 2))
plot (survfit (Surv(pData(vdx)$t.dmfs , pData(vdx)$e.dmfs) ~
  pData(vdx)$er) , col = c("cyan" , "salmon"))

plot (survfit(Surv(pData(vdx)$t.dmfs , pData(vdx)$e.dmfs) ~
  pData (vdx)$grade) , col = c("blue" , "yellow" , "orange"))

survdiff(Surv(pData(vdx)$t.dmfs , pData(vdx)$e.dmfs) ~
  pData (vdx)$er)

survdiff(Surv(pData(vdx)$t.dmfs , pData(vdx)$e.dmfs) ~
  pData(vdx)$grade)
```

