---
title: "Importing publicly available data"
author: "Your Name Here"
date: '15th Feb. 2016'
output: html_document
---

```{r}
remotefile <- 'ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE1nnn/GSE1729/matrix/GSE1729_series_matrix.txt.gz'
destfile <- "data/GSE1729_series_matrix.txt.gz"
if(!file.exists(destfile)) download.file(remotefile, destfile)
mydata <- getGEO(filename="data/GSE1729_series_matrix.txt.gz")
mydata
```


******
## Q. What scale are the expression values recorded on? Is this an appropriate scale for analysis / visualisation?

******


```{r}
exprs(mydata)[1:5,1:5]
summary(exprs(mydata)[,1:5])
```


```{r}
exprs(mydata)[1,]
fData(mydata)[1,1:8]
```



## Example 1.1 Quality assessment of the downloaded data



```{r}
boxplot(log2(exprs(mydata)))
```



******
## Q. Generate an arrayQualityMetrics report for the dataset. Do any arrays seem to be poor quality?

The report should be generated in a folder called `arrayQualityMetrics report for mydata` (the name of the directory can be configured).

N.B the `force=TRUE` argument will make sure that the output gets written if the specified directory name already exists
******

```{r}
library(arrayQualityMetrics)
arrayQualityMetrics(mydata, force=TRUE)
```


```{r}
a <- matrix(rnorm(100),ncol=10)
colnames(a) <- LETTERS[1:10]
a[,1]
a[,-1]
```


```{r}
a[,c(1,10)]
a[,-c(1,10)]
```



```{r}
subset <- mydata[,1:4]
dim(exprs(subset))
dim(pData(subset))
```


```{r}
subset <- mydata[,-c(1,4)]
subset
```



******
## Q. Now create a new dataset with the poor quality arrays removed

******

```{r}

```



## Example 1.2 - Dealing with gene annotation


```{r}
fData(mydata)[1:5,1:5]
colnames(fData(mydata))
all(rownames(fData(mydata)) == rownames(exprs(mydata)))
```


```{r}
features <- fData(mydata)
```



```{r}
colnames(features)
```


******

## Exercise

- How many unique Entrez gene IDs are there?
- How many features do not have an associated Entrez ID
- Which Entrez ID has the most probes?
- What probes are used for the gene *TP53*

******

```{r}

```


##Example: Retrieving data for a particular gene



******
Q. Use the following code to understand the difference between `grep`, `match` and `==`

******

```{r}
features[grep("TP53", features$`Gene Symbol`),]
features[which(features$`Gene Symbol` == "TP53"),]
features[match("TP53", features$`Gene Symbol`),]
```


```{r}
rows <- which(features$`Gene Symbol` == "TP53")
rows
```


```{r}
E <- log2(exprs(mydata))
E[1274,]
E[10723,]
```


```{r}
par(mfrow=c(1,2))
plot(E[1274,], xlab="Array Index ",
     col="steelblue",pch=16,
     ylab="Normalised Expression",main="201746_at")

plot(E[10723,], xlab="Array Index ",
     col="steelblue",pch=16,
     ylab="Normalised Expression",main="211300_s_at")

```


```{r}
par(mfrow=c(1,1))
plot(E[1274,],E[10723,],xlab="201746_at",ylab="211300_s_at",col="steelblue",pch=16)
cor(E[1274,],E[10723,])
```


******
Q. Use the `IQR` function to calculate the inter-quartle range of both probes. Which would be the best probe to report in an analysis?
******

```{r}
IQR(E[1274,])
IQR(E[10723,])
```



## Example 2. Dealing with large cohorts

```{r}
remotefile <- 'ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE18nnn/GSE18088/matrix/GSE18088_series_matrix.txt.gz'
destfile <- "data/GSE18088_series_matrix.txt.gz"
if(!file.exists(destfile)) download.file(remotefile, destfile)
cohort <- getGEO(filename="data/GSE18088_series_matrix.txt.gz")
cohort
```


```{r}
head(exprs(cohort)[,1:5])
head(fData(cohort)[,1:5])
```


```{r}
pData(cohort)[1:5,1:5]
colnames(pData(cohort))
```

```{r}
pd <- pData(cohort)
View(pd)
```


```{r}
pd[1:4,10:16]
```

******

Q. How many males and females are there in the study?
Q. How many of the patients relapsed?

******

```{r}
```



```{r}
rows <- which(fData(cohort)$"Gene Symbol" == "XIST")
rows
```

```{r}
subset <- exprs(cohort)[rows,]
dim(subset)
plot(subset[1,])
plot(density(subset[1,]))
```


```{r}
boxplot(subset[1,]~pd$characteristics_ch1.1)
```

***N.B. we haven't discussed looping in R yet, so here is the long-hand way to produce these plots***

```{r}
par(mfrow=c(2,4))
boxplot(subset[1,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
boxplot(subset[2,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
boxplot(subset[3,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
boxplot(subset[4,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
boxplot(subset[5,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
boxplot(subset[6,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
boxplot(subset[7,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
boxplot(subset[8,]~pd$characteristics_ch1.1,las=2,main=fData(cohort)[rows[1],1])
```

## Example: Extra manipulation of the clinical data


******
Q. What is the age distribution of the samples?

******

```{r}
clinvars <- pd[,10:16]
gsub("age at diagnosis, years: ","",clinvars$characteristics_ch1.4)
```


```{r}
clinvars$characteristics_ch1.4 <- as.numeric(gsub("age at diagnosis, years: ","",clinvars$characteristics_ch1.4))
hist(clinvars$characteristics_ch1.4)
```

******
Q. Clean-up the gender column so that the values are either `male` or `female`.

******

```{r}

```




